Unrecognized States: Theory and Cases

### 2017-07-01 theory.tex
remove appendix

Despite the preferences of the international community for peace, the most common outcome of secessionist conflicts in the post-WWII period has been reunification with the home state via outright military reconquest. By contrast, reunification through negotiated settlement has been very rare. We will explore potential explanations for this seeming inability of the international community to achieve its aims in Sections~\ref{sec:alt} and \ref{sec:ext}. First, we turn to the outcome that we find the most puzzling: perpetual unrecognized statehood.

\subsection{Analysis of the ``Status Quo'' Equilibrium} 
\label{sec:mainproof}
Unrecognized states are frequently viewed as temporary phenomena or as non-equilibrium outcomes attributable to players' misperceptions of the strategic situation, or their fundamental irrationality. Our central result shows that unrecognized statehood can be an equilibrium outcome capable of being sustained in perpetuity by fully rational, perfectly informed actors. This is true even when there is no actor that prefers unrecognized statehood as a first-best outcome.

@@

The proofs of the lemmas are in the Appendix.

@@

The proof of Proposition 2 is in the Appendix.

@@

\pagebreak
\appendix
\section{Appendix}

{\bf Proof of Lemma 1} \\
Recall that the upper case notation represents payoffs gross of investments by the outside players. Without loss of generality, consider the incentives to deviate for player $s$ in period $t$ given that the other three players play their equilibrium actions. Given the stationary equilibrium actions, playing Status Quo will lead to $Q_s^t \ \forall t$, so the continuation value is $\frac{Q_s^t}{1-\delta}$. Player $s$'s continuation value from the one-shot deviation to Cede, which leads to an absorbing state with payoff $L_s^t$ in each period, is $\frac{L_s^t}{1-\delta}$. The one-shot deviation to fight results in the war lottery, with a cost of $-\zeta_s$, $W_s$ forever with probability $p$, and $L_s$ forever with probability $1-p$. Thus the continuation value is $-\zeta_s +\frac{W_s^t(p)+ L_s^t(1-p)}{1-\delta}$.

Given that the continuation value from Status Quo is higher than both that from playing Cede and Fight, player $s$ will play Status Quo. The argument for player $g$ is symmetric. \hfill $\blacksquare$
\\
\\
\noindent {\bf Proof of Lemma 2} \\
Begin by assuming Restriction \ref{res:new}a holds. In the status quo, we have $U_p^{t,SQ}= -R_p^{SQ}$ and $U_c^{t,SQ}= -R_c^{SQ}$. If the outcome is reunification, payoffs in this absorbing state in the period in which the investment is made are $U_p^{t,RU}= -\alpha -R_p^{RU}$ and $U_c^{t,RU}= \beta -R_c^{RU}$ where $SQ$ and $RU$ distinguish investments under the status quo and reunification scenarios respectively.

Since reunification is the patron's least preferred state, $R_p^{RU} =0$; that is, the patron is not willing to invest to provoke reunification. But the patron \textit{would} be willing to invest the amount it would lose from reunification to maintain the Status Quo: that is, $R_p^{SQ} = \frac{\alpha}{1 -\de}$.

For player $c$ the difference in continuation values from switching from the status quo to reunification is $\frac{\beta + R_c^{SQ}}{1 -\de} -R_c^{RU}$. So $c$ is willing to invest up to $\frac{\beta+ R_c^{SQ}}{1 -\de}$ toward reunification---or preventing the status quo.

Player $c$ would be able to induce the secessionists to play Cede---e.g. counter an addition to $Q_s^{t-1}$ with an investment to $L_s^{t-1}$---as long as $R_c^{RU} = \frac{\beta + R_c^{SQ}}{1 -\de} > Q_s^{t-1} -\mu + R_p^t - L_s^{t-1}$.\footnote{Here we assume that the payoff relationship at $t$ is such that ceding is better than fighting. If this relationship is reversed, similar analysis and results hold.} For this to work, the investment required of player $c$ must be no more than the gain player $c$ receives from investing.

On the other hand, if $\frac{\beta + R_c^{SQ}}{1 -\de} \leq Q_s^{t-1} -\mu + R_p^t - L_s^{t-1}$, the patron's investment creates a large enough difference in the secessionist's payoffs from the status quo versus ceding (net of player $c$'s investment) so that player $c$ will not find contesting the status quo to be in its interest. Whenever the status quo will be the outcome in this way, $c$ maximizes its utility by choice of $R_c^{SQ} =0$.

Restriction (\ref{res:1}) combined with equilibrium play implies that  $Q_s^{t-1} > L_s^{t-1}$. Combining this with $R_c^{SQ} =0$, a tighter condition for preventing player $c$ from inducing reunification is $\frac{\beta}{1 -\de} + \mu \leq R_p^t$.

Since the patron is willing to invest up to $\frac{\alpha}{1 -\de}$, $\frac{\beta}{1 -\de} + \mu \leq \frac{\alpha}{1 -\de}$ is a sufficient condition to prevent $c$ from inducing the \emph{reunification} outcome.

If Restriction \ref{res:new}a is violated, as the secessionists' status quo payoffs degrade from period to period, they will fall below $\omega_s^t$ before they fall below $L_s^t$. If no investment is made in $Q_s^{t-1}$, the secessionists will choose the war outcome. Restriction \ref{res:new}b guarantees that the patron will make the investment $R_p^{t'}  = \frac{\omega_s^{t-1}}{1-\de} +\mu - \frac{Q_s^{t-1}}{1-\de} \leq \mu$ to prevent the secessionists choosing war since this $R_p^{t'}$ is strictly less than the patron's expected loss from the war outcome. Note that the patron's actual investment will be the maximum of the investment required to prevent the war and ceding outcomes. \hfill $\blacksquare$
\\
\\
\noindent {\bf Proof of Lemma 3} \\
Again, in the status quo, we have $U_p^{t,SQ}= -R_p^{SQ}$ and $U_c^{t,SQ}= -R_c^{SQ}$. If the outcome is recognition of the secessionists, per-period payoffs in this absorbing state become $U_p^{t,REC}= \lambda -R_p^{REC}$ and $U_c^{t,REC}= -\nu -R_c^{REC}$ where $SQ$ and $REC$ distinguish investments under the status quo and recognition scenarios respectively.

Thus the difference in continuation values from switching from the status quo to recognition for player $p$ is $\frac{\lambda +R_p^{SQ}}{1 -\de} -R_p^{REC}$. For player $c$ it is $\frac{-\nu + R_c^{SQ}}{1 -\de}-R_c^{REC}$. $R_c^{REC} =0$ since the international community will not invest to create its least preferred outcome. So $c$ is willing to invest up to $\nu$ per period toward maintaining the status quo, and $p$ is willing to invest up to $\frac{\lambda +R_p^{SQ}}{1 -\de}$ toward recognition. We have shown that the largest possible investments by $p$ in the status quo are $\frac{\beta}{1 -\de}$ in $t=1$ and $\mu$ in each period thereafter, so the largest this difference can be for $p$ is $\frac{\lambda + \beta + \mu}{1 -\de} -\mu$.

If $\frac{\nu}{1-\de} < L_g^{t-1} + R_p^t - Q_g^{t-1}$, the patron's investment to augment $L_g^{t-1}$ creates a large enough difference in the secessionists' payoffs of ceding versus the status quo (net of player $c$'s investment) so that player $c$ will not find it worthwhile to re-establish the status quo outcome.

On the other hand, player $c$ would be willing to counter whatever investment player $p$ makes to try to induce recognition---e.g. counter an addition to $L_g^{t-1}$ with an investment to augment $Q_g^{t-1}$---as long as $\frac{\nu}{1-\de} \geq L_g^{t-1} + R_p^t - Q_g^{t-1}$.\footnote{As in the proof of Lemma 2, we have assumed that the payoff relationship at $t$ is such that ceding is better than fighting. If this relationship is reversed, similar analysis and results hold.}

Restriction (\ref{res:1}) combined with equilibrium play implies that  $Q_s^{t-1} > L_s^{t-1}$ so that a tighter condition for preventing player $p$ from inducing recognition is $\frac{\nu}{1 -\de} \geq R_p^t$.

Since the patron is willing to invest up to $\frac{\lambda + \beta + \mu}{1 -\de} -\mu$, when $\frac{\nu}{1 -\de} \geq \frac{\lambda + \beta + \mu}{1 -\de} -\mu$, player $c$'s willingness to counter the patron's investments is sufficient to prevent the \emph{recognition} outcome. A slightly weaker version of this inequality is stated as Restriction (\ref{res:4}) so that this deterrence dynamic is guaranteed to operate through the war lottery as well, as will be seen in Lemma 4. \hfill $\blacksquare$
\\
\\
\noindent {\bf Proof of Lemma 4} \\
The difference in continuation values from switching from the status quo to taking the war lottery for player $p$ is $\frac{\lambda p_s - \alpha (1-p_s)}{1 -\de} +R_p^{SQ} -R_p^{WAR}$. For player $c$ it is $\frac{-\nu p_s + \beta (1-p_s)}{1 -\de} + R_c^{SQ} -R_c^{WAR}$. As noted above, $R_c^{SQ} =0$. So $c$ is willing to invest up to $\frac{\nu p_s - \beta (1-p_s)}{1 -\de}$ toward preventing fighting.

The patron is willing to invest up to $\frac{\lambda p_s - \alpha (1-p_s)}{1 -\de} +R_p^{SQ}$ toward instigating fighting when this quantity is positive, and will not invest when the quantity is non-positive. Thus we must only examine the case where this quantity is positive.

We will again set $R_p^{SQ}$ at its largest value of $\frac{\mu + \beta}{1-\de} -\mu$ as this is most restrictive. In this only case of interest, we have $\frac{\lambda p_s - \alpha (1-p_s) + \mu + \beta}{1 - \de} -\mu> 0$ for an upper bound on the amount the patron is willing to invest to instigate fighting.

If $\frac{\nu p_s - \beta (1-p_s)}{1 -\de}  < \omega_s^t\left(R_p^t\right) - Q_s^t$, the patron's investment in $\omega_s^t$ creates a large enough difference between the payoffs from fighting and the status quo so that player $c$ will not find it worthwhile to re-establish the status quo outcome.

On the other hand, player $c$ would be willing to counter whatever investment player $p$ makes to try to induce fighting---e.g. counter an addition to $\omega_s^t$ with an investment to augment $Q_s^{t-1}$---as long as $\frac{\nu p_s - \beta (1-p_s)}{1 -\de}  \geq \omega_s^t\left(R_p^t\right) - Q_s^t$.

This can be expanded as $\frac{\nu p_s - \beta (1-p_s)}{1 -\de}  \geq -\zeta_{s}(1-\de) + (W_s^{t-1}+R_p^t)p_s + L_s^{t-1}(1-p_s) - Q_s^{t-1} +\mu = \omega_s^{t-1} - Q_s^{t-1} +R_p^t p_s + \mu$.

Restriction (\ref{res:1}) combined with equilibrium play implies that  $Q_s^{t-1} \geq \omega_s^{t-1}$ so a tighter condition for preventing player $p$ from inducing fighting is $\frac{\nu p_s - \beta (1-p_s)}{1 -\de} - \mu \geq R_p^t p_s$.

Since $R_p^t \leq \frac{\lambda p_s - \alpha (1-p_s) + \mu + \beta}{1 - \de} -\mu$, if we can show $\frac{\nu p_s - \beta (1-p_s)}{1 -\de}  - \mu \geq p_s \frac{\lambda p_s - \alpha (1-p_s) + \mu + \beta}{1 - \de} - p_s \mu$, we will have the guarantee that player $c$ will counterinvest after any effort by the patron to instigate fighting, thereby making $p$'s optimal investment in fighting zero.

Multiplying Restriction (\ref{res:3}) by $-(1-p_s)$, we have $\frac{-\beta \cdot (1-p_s)}{1-\de} -\mu (1-p_s) \geq \frac{-\alpha \cdot (1-p_s)}{1-\de}$. Using this with Restriction (\ref{res:4}) multiplied by $\frac{1}{1-\de}$, we have $\frac{\nu p_s - \beta (1-p_s)}{1-\de} -\mu \geq \frac{\lambda p_s - \alpha (1-p_s) + \mu + \beta}{1-\de} -p_s \mu$. Since it is assumed that the war lottery probabilities are non-degenerate, we also have $\frac{\lambda p_s - \alpha (1-p_s) + \mu + \beta}{1-\de} -p_s \mu > p_s \frac{\lambda p_s - \alpha (1-p_s) + \mu + \beta}{1-\de} -p_s \mu$. \hfill $\blacksquare$
\\
\\
{\bf Proof of Proposition 2} \\
Suppose that sanctions are applied by player $c$ in period $t+1$ after a Status Quo equilibrium such as the one supported by the construction in Proposition 1 has been established in period $t$. As the only alteration is a reduction in $Q_s^t$ by the amount of the sanctions, it is immediate that if player $p$ is willing and able to supplement player $s$'s status quo payoffs to counteract sanctions, then player $s$'s incentives to play Status Quo are not disturbed. As in Proposition 1, it is not necessary that the patron make these payments immediately if player $s$'s utility from playing Status Quo is high enough relative to its other options, but the degradation of the status quo payoffs will eventually make these payments necessary in order to sustain a Status Quo equilibrium.

Therefore if sanctions are to disrupt the Status Quo equilibrium and potentially lead to ceding by the secessionists, it must be that the patron is either not able or not willing to make the investment required to maintain the status quo under player $c$'s sanctions (Condition 1).

Assume then that Condition 1 holds so that the patron makes no investments in the Status Quo payoffs of player $s$. We also need the patron to withhold investments toward encouraging the secessionists to fight. Lemma 4 of Proposition 1 is not sufficient to guarantee this behavior since no investments are being made in the Status Quo. In addition, the reference outcome changes: players $p$ and $c$ must now compare their payoffs in the war scenario to the reunification outcome. To be precise, the least stringent condition is $m$ periods of Status Quo followed by the Reunification outcome, where $m$ is the maximal number of periods that it takes for the sanctions to degrade the status quo payoffs below the level of the payoffs from ceding for player $s$.

Player $p$'s continuation value when the status quo is played for $m-1$ periods before ceding is $0 - \frac{\de^m}{1-\de} \alpha$, whereas the continuation value from war is $\frac{\lambda p_s - \alpha (1-p_s)}{1-\de}$. For player $c$, it is $0 + \frac{\de^m}{1-\de} \beta$ versus $\frac{-\nu p_s + \beta (1 - p_s)}{1-\de}$. Thus the condition for $p$'s willingness to instigate fighting to outweigh $c$'s willingness to deter fighting is
\[
  \textstyle \frac{\lambda p_s - \alpha (1-p_s)}{1-\de} + \frac{\de^m}{1-\de} \alpha > \frac{-\nu p_s + \beta (1-p_s)}{1-\de} -  \frac{\de^m}{1-\de} \beta
\]

If this condition holds, then there will be no investments following sanctions. But it is still not guaranteed that the secessionists will choose to cede. Status Quo is no longer a long-run equilibrium strategy, but even without outside investments, it is possible that Fight could be a more attractive action than Cede.

The status quo payoffs will degrade period by period in the absence of investments. Whether the secessionists choose Cede or Fight depends on which strategy yields the higher continuation value. In order to avoid conflict, sanctions must cause player $s$'s continuation value from playing Status Quo to fall below its continuation value from playing Cede before it falls below the continuation value from playing Fight. \hfill $\blacksquare$

%Several possibilities exist. First, the sanctions could degrade the payoffs to ceding more than they degrade the status quo payoffs. If $S_L > \mu + S_Q$ where $S_L$ and $S_Q$ are the per-period reduction in the payoffs to ceding and maintaining the Status Quo, respectively, both payoffs will degrade until eventually player $s$ will choose \emph{Fight}. This occurs because the continuation value from \emph{Cede} will remain below that from playing \emph{Status Quo}, but the sanctions only have a partial effect in reducing the \emph{Fight} continuation value through the war lottery, while their entire impact is felt on the \emph{Status Quo} continuation value. The gap between the Status Quo and Fight continuation values that was positive without sanctions will eventually become negative, leading to conflict.

%Thus we must have $\mu + S_Q > S_L$ if there is any chance that the secessionists will choose to cede.

%\newpage

%\theendnotes

>>>>

Despite the preferences of the international community for peace, the most common outcome of secessionist conflicts in the post-WWII period has been reunification with the home state via outright military reconquest. By contrast, reunification through negotiated settlement has been very rare. We will explore potential explanations for this seeming inability of the international community to achieve its aims in Section~\ref{sec:alt}. First, we turn to the outcome that we find the most puzzling: perpetual unrecognized statehood.

\subsection{Analysis of the ``Status Quo'' Equilibrium} 
\label{sec:mainproof}
Unrecognized states are frequently viewed as temporary phenomena or as non-equilibrium outcomes attributable to players' misperceptions of the strategic situation, or their fundamental irrationality. Our central result shows that unrecognized statehood can be an equilibrium outcome capable of being sustained in perpetuity by fully rational, perfectly informed actors.

@@

The proofs of the lemmas can be found in Buzard, Graham and Horne (2017).

@@

The proof of Proposition 2 can be found in Buzard, Graham and Horne (2017).

@@
